---
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.6.0
  kernelspec:
    display_name: R 4.0.2
    language: R
    name: ir4.0.2
---

<!-- #region slideshow={"slide_type": "slide"} -->
# Box-Cox transformation Demo
Using Random data, we generate a polynomial dataset with $x \in [0.1 , 10]$. 
Using the boxcox R function, we estimate the best box cox transform to get a straight data set. 

With the  `?boxcox` command you will see what other parameters you can give the boxcox function. 
More information about Box-Cox transforms by Peng-Fei Li: https://www.ime.usp.br/~abe/lista/pdfm9cJKUmFZp.pdf
The original script (without randomization) can be found at https://www.statology.org/box-cox-transformation-in-r/
<!-- #endregion -->

```{r slideshow={'slide_type': 'fragment'}}
require(MASS)
#require(ggplot2)
```

```{r slideshow={'slide_type': 'fragment'}}
#?runif
#?rdisc
n = 1000
r = rnorm(1,2,2)
x=runif(n,0.1,10)
y = 20+ x^r *(1+ rnorm(n,0,.3))
print(paste("polynomial: y = 20 +x^" , round(r,4), "+eps"))
```

```{r slideshow={'slide_type': 'subslide'}}
#find optimal lambda for Box-Cox transformation 
bc <- boxcox(y ~ x, lambda = seq(-3,3,0.1))
(lambda <- bc$x[which.max(bc$y)])
#fit new linear regression model using the Box-Cox transformation
new_model <- lm(((y^lambda-1)/lambda) ~ x)
```

<!-- #region slideshow={"slide_type": "slide"} -->
### Fit linear regression models
<!-- #endregion -->

```{r slideshow={'slide_type': 'fragment'}}
model <- lm(y~x)
#fit new linear regression model using the Box-Cox transformation
new_model <- lm(((y^lambda-1)/lambda) ~ x)
#define plotting area
op <- par(pty = "s", mfrow = c(2, 2))

#Q-Q plot for original model
qqnorm(model$residuals)
qqline(model$residuals)

#Q-Q plot for Box-Cox transformed model
qqnorm(new_model$residuals)
qqline(new_model$residuals)
plot(x,y,pch = ".")#, col = c(50,50,50,alpha = .9))
title("original scatterplot and lm fit")
abline(model)
plot(x, (y^lambda-1)/lambda, pch = ".") #, col =c(50,50,50,alpha = .5))
title("transformed scatterplot and lm fit")
abline(new_model)
#display all 4 plots
par(op)
```

<!-- #region slideshow={"slide_type": "subslide"} -->
If we are lucky (with the random exponent), the transformed data shows a much straighter QQ plot, showing that the residuals are more normally distributed, and the right hand side scatterplot has a more equal variation and is straighter around the line. 

If the exponent is larger than 1, it works well. if the exponent is 03 or -3, it doesnt work well, i.e. the QQ plot on the right hand side shows no less deviation from a straight line than the untransformed one on the LHS, and the Regression fit plot on the RHS shows the transformed data is not better fitted to a straight line than the untransformed data on the LHS.  
Run the above cells a few times to convince yourself of this. 
<!-- #endregion -->

<!-- #region slideshow={"slide_type": "slide"} -->
### A summary of estimations of the original and transformed models:
<!-- #endregion -->

```{r slideshow={'slide_type': '-'}}
summary(model)
summary(new_model)
```

<!-- #region slideshow={"slide_type": "slide"} -->
## Other data distributions
<!-- #endregion -->

<!-- #region slideshow={"slide_type": "subslide"} -->
### Poisson distributed data: (e.g. # of potholes in a km of road)
<!-- #endregion -->

```{r slideshow={'slide_type': 'fragment'}}
#find optimal lambda for Box-Cox transformation of Poisson data
require(purrr)
m <- runif(1,0.1,10)
df <- data.frame( t=seq(from=1, to = 5, length.out= n))
df$z <- map_dbl(df$t, {function(t) { rpois(1,t) +.1} })
op <- par(pty = "s", mfrow = c(1, 2))  #2 plots side by side
hist(df$z/df$t,prob =T)
lines(density(df$z))
lines(density(df$z, adjust=2), lty="dotted")
message(paste("z = Poisson(t)+.01  
   ( +.01 to be positive: Box Cox transforms only work for positive variables)"))
#print(bcz)
bcz <- boxcox(df$z ~ df$t, lambda = seq(-3,3,0.1))
lambdaz <- bcz$x[which.max(bcz$y)]
message('best exponent seems to be' )
message(lambdaz)
par(op)
```

```{r slideshow={'slide_type': 'subslide'}}
modelz <- lm(df$z ~ df$t)
#fit new linear regression model using the Box-Cox transformation
#print( (df$z^lambdaz-1)/lambdaz)

new_modelz <- lm(((df$z^lambdaz-1)/lambdaz) ~ df$t)

#define plotting area
op <- par(pty = "s", mfrow = c(2, 2))

#Q-Q plot for original model
qqnorm(modelz$residuals)
qqline(modelz$residuals)

#Q-Q plot for Box-Cox transformed model
qqnorm(new_modelz$residuals)
qqline(new_modelz$residuals)
plot(df$t,df$z,pch = ".")
title("original scatterplot and lm fit")
abline(modelz)
plot(df$t, (df$z^lambdaz-1)/lambdaz,pch = ".")
title("transformed scatterplot and lm fit")
abline(new_modelz)
#display all 4 plots
par(op)
```

```{r slideshow={'slide_type': 'subslide'}}
summary(modelz)
summary(new_modelz)
```

<!-- #region slideshow={"slide_type": "slide"} -->
### Exponentially distributed data (# of km between potholes on road)
<!-- #endregion -->

```{r slideshow={'slide_type': 'fragment'}}
#find optimal lambda for Box-Cox transformation of exponential data


df$e <- map_dbl(df$t, {function(t) { rexp(1,t) +.01} })
op <- par(pty = "s", mfrow = c(1, 2))
hist(df$e/df$t,prob =T)
lines(density(df$e))
lines(density(df$e, adjust=2), lty="dotted")
message(paste("e = exp(t)+.01  
   ( +.01 to be positive: Box Cox transforms only work for positive variables)"))

bce <- boxcox(df$e ~ df$t, lambda = seq(-3,3,0.1))
lambdae <- bce$x[which.max(bce$y)]
message('best exponent seems to be' )
message(lambdae)
par(op)
```

```{r slideshow={'slide_type': 'subslide'}}
modele <- lm(df$e ~ df$t)
#fit new linear regression model using the Box-Cox transformation

new_modele <- lm(((df$e^lambdae-1)/lambdae) ~ df$t)

#define plotting area
op <- par(pty = "s", mfrow = c(2, 2))

#Q-Q plot for original model
qqnorm(modele$residuals)
qqline(modele$residuals)

#Q-Q plot for Box-Cox transformed model
qqnorm(new_modele$residuals)
qqline(new_modele$residuals)
plot(df$t,df$e,pch = ".")
title("original scatterplot and lm fit")
abline(modele)
plot(df$t, (df$e^lambdae-1)/lambdae,pch = ".")
title("transformed scatterplot and lm fit")
abline(new_modele)
#display all 4 plots
par(op)
```

```{r slideshow={'slide_type': 'subslide'}}
summary(modele)
summary(new_modele)
```
